Avoiding the Global Sort: \break A Faster Contour Tree Algorithm \footnote{The full updated version of this paper is available on the arXiv \cite{rs-mvrms-14}}

### 2016-07-08 paint-contour-tree.tex
typos corrected

Essentially, the ``run time per critical point" is the height/depth of the point in the contour tree.
This bound immediately yields a run time of $O(t\log D + t\alpha(t) + N)$,
where $D$ is the diameter of the contour tree. 
This is a significant asymptotic improvement for short and fat contour trees. For example, if the tree is balanced,
then we get a bound of $O(t\log\log t)$.
Even if $T$ contains a long path of length $O(t/\log t)$, but is otherwise short, we get the improved bound of $O(t\log\log t)$.

@@

We denote the contour tree of $\MM$
by $\reeb(\MM)$. The corresponding node and edge sets are denoted as $\cV(\cdot)$ and $\cE(\cdot)$.
It is not immediately obvious that this graph is a tree, but alternate definitions of the contour tree
in~\cite{csa-cctad-03} imply this is a tree. 
Since this tree has height values associated with the vertices, we can talk about up-degrees and down-degrees in $\reeb(\MM)$.
%
Similar to \cite{kobps-ctsssit-97} (among others), multi-saddles are treated as a set of ordinary saddles, which can be realized 
via vertex unfolding (which can increase surface complexity if multi-saddle degrees are allowed to be super-constant).
Therefore, to simplify the presentation, for the remainder of the paper up and down-degrees are at most $2$, and total degree is at most $3$.
%
%We assume there are no ``multi-saddles", so up and down-degrees are at most $2$, and the total degree is at most $3$. 
%This is again a standard assumption in topological algorithms and can be achieved by vertex unfolding (Section 6.3 in~\cite{HaEd10}).

@@

The paths in $P(T)$ naturally define a tree\footnote{Please excuse the 
overloading of the term 'tree', it is the most natural term to use here.} of their own.  Specifically, in the original 
tree $T$ contract each path down to its root.  Call the resulting tree the \emph{shrub} of $T$ corresponding to 
the path decomposition $P(T)$. Abusing notation, we simply use $P(T)$ to denote the shrub.
As a result, we use terms like `parent', `child', `sibling', etc. for paths as well.
The shrub gives a handle on the heaps of a path.
We use $b(p)$ to denote the \emph{base} of the path, which is vertex in $p$ closest
to root of $T$. We use $\ell(p)$ to denote the leaf in $p$.
We use $H_p$ to denote the $H_{b(p)}$.
% and denote it by $\pathTree(T)$.

\begin{lemma}
\label{lem:adjacent}
 Let $p$ be any path in $P(T)$ and let $\{q_1, \dots q_k\}$ be the children on $p$.
Then $|H_{\ell(p)}| + \sum_{i=1}^k |H_{q_i}| \leq |H_p|+2|p|$. 
\end{lemma}

@@

Let $\cL$ be the set of short paths, $\cL'$ be the set of supported short paths, and $\cH$ be the set of tall paths
given by $\pmax(T)$.
We now construct the shrub of unsupported short paths. Consider $p \in \cL \setminus \cL'$,
and traverse the chain of ancestors from $p$. Eventually, we must reach another short path $q$.
(If not, we have reached the root $r$ of $\pmax(T)$. Hence, $p$ is supported.)
Insert edge from $p$ to $q$, so $q$ is the parent of $p$ in $\cU$. This construction leads
to the shrub forest of $\cL \setminus \cL'$, where all the roots are supported short paths, 
and the remaining nodes are the unsupported short paths.

@@

We split the summation into tall, short, and unsupported short paths.
\begin{eqnarray*} 
\sum_{p \in \cL} \sum_{v \in p} \log |H_v| & = & \sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| + \sum_{p \in \cL'} \sum_{v \in p} \log |H_v| + \sum_{p \in \cH} \sum_{v \in p} \log |H_v|
%& = & \sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| + O(\sum_{p \in \cL'} |p|\log |H_p|) = \sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| + O(\sum_{p \in \cL'} |H_p|\log |H_p|)
\end{eqnarray*}
The last term can be bounded by $O(\sum_{p \in \pmax(T)} |p|\log |p|)$, by \Lem{pathBounds}.
The second term can be bounded by $O(\sum_{p \in \cL'} |H_p| \log |H_p|)$, by \Lem{pathBounds} again.
The following claim shows that this in turn is at most the last term.

@@

\begin{proof} Let $\cU$ denote a connected component of the shrub forest.
We have $\sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| \allowbreak \leq \sum_{\cU} \sum_{p \in \cU} \sum_{v \in p} \log |H_v|$.
By \Lem{cu-root}, $|H_v| = O(|H_r|)$, where $r$ is the root of $\cU$.
Furthermore, $\sum_{p \in \cU} |p| = O(|H_r|)$. 
We have $\sum_{p \in \cU} \sum_{v \in p} \log |H_v| = O((\log |H_r|) \sum_{p \in \cU} |p|) = O(|H_r|\log |H_r|)$.
We sum this over all $\cU$ in the shrub forest, and note that roots in the shrub forest are supported short paths.
\end{proof}

@@

We define a \emph{residue} $R_p$ for each $p \in \cU$. Suppose $p$ has children
$q_1, q_2, \ldots, q_k$ in $\cU$. Then $R_p = |\redH_p| - \sum_i |\redH_{q_i}|$. 
By definition, $|\redH_p| = \sum_{q \in \cU_p} R_p$. 
Note that $R_p$ can be negative. 
Now, define $R^+_p = \max(R_p,0)$, and set $W_p = \sum_{q \in \cU_p} R^+_p$.
Observe that $W_p \geq |\redH_p|$. We also get an approximate converse.

@@

\begin{proof} We write $W_p - |\redH_p| = \sum_{q \in \cU_p} R^+_q - R_q$ $= -\sum_{q \in \cU_p: R_q < 0} R_q$.
Consider $q \in \cU_p$ and denote the children in $\cU_p$ by $q'_1, q'_2, \ldots$.
Note that $R_q$ is negative exactly when $|\redH_q| < \sum_i |\redH_{q'_i}|$.
Traverse $\pmax(T)$ from $q'_i$ to $q$.
Other than $q$, all other nodes encountered are in the tall support chain of $q'_i$ and hence
do not affect its reduced heap.
The vertices of $\redH_{q'_i}$ that are deleted
are exactly those present in the path $q$. 
Any vertex in $q$ can be deleted from at most two of the reduced heaps (of the children of $q$ in $\cU_p$), since 
theses reduced heaps do not have an ancestor-descendant relationship.
Therefore when $R_q$ is negative, it is at most by $2|q|$.
We sum over all $q$ to complete the proof.
\end{proof}

>>>>

Essentially, the ``run time per critical point" is determined by the height/depth of the point in the contour tree.
This bound immediately yields a run time of $O(t\log D + t\alpha(t) + N)$,
where $D$ is the diameter of the contour tree. 
This is a significant asymptotic improvement for short and fat contour trees. For example, if the tree is balanced,
then we get a bound of $O(t\log\log t)$.
Even if $T$ contains a long path of length $O(t/\log t)$, but is otherwise short, we get the improved bound of $O(t\log\log t)$.

@@

We denote the contour tree of $\MM$
by $\reeb(\MM)$. The corresponding node and edge sets are denoted as $\cV(\cdot)$ and $\cE(\cdot)$.
It is not immediately obvious that this graph is a tree, but alternative definitions of the contour tree
in~\cite{csa-cctad-03} imply this is a tree. 
Since this tree has height values associated with the vertices, we can talk about up-degrees and down-degrees in $\reeb(\MM)$.
%
Similar to \cite{kobps-ctsssit-97} (among others), multi-saddles are treated as a set of ordinary saddles, which can be realized 
via vertex unfolding (which can increase surface complexity if multi-saddle degrees are allowed to be super-constant).
Therefore, to simplify the presentation, for the remainder of the paper up and down-degrees are at most $2$, and total degree is at most $3$.
%
%We assume there are no ``multi-saddles", so up and down-degrees are at most $2$, and the total degree is at most $3$. 
%This is again a standard assumption in topological algorithms and can be achieved by vertex unfolding (Section 6.3 in~\cite{HaEd10}).

@@

The paths in $P(T)$ naturally define a tree\footnote{Please excuse the 
overloading of the term 'tree', it is the most natural term to use here.} of their own.  Specifically, in the original 
tree $T$ contract each path down to its root.  Call the resulting tree the \emph{shrub} of $T$ corresponding to 
the path decomposition $P(T)$. Abusing notation, we simply use $P(T)$ to denote the shrub.
As a result, we use terms like `parent', `child', `sibling', etc.\ for paths as well.
The shrub gives a handle on the heaps of a path.
We use $b(p)$ to denote the \emph{base} of the path, which is the vertex in $p$ closest
to root of $T$. We use $\ell(p)$ to denote the leaf in $p$.
We use $H_p$ to denote $H_{b(p)}$.
% and denote it by $\pathTree(T)$.

\begin{lemma}
\label{lem:adjacent}
 Let $p$ be any path in $P(T)$ and let $\{q_1, \dots q_k\}$ be the children of $p$.
Then $|H_{\ell(p)}| + \sum_{i=1}^k |H_{q_i}| \leq |H_p|+2|p|$. 
\end{lemma}

@@

Let $\cL$ be the set of short paths, $\cL'$ be the set of supported short paths, and $\cH$ be the set of tall paths
given by $\pmax(T)$.
We now construct the shrub of unsupported short paths. Consider $p \in \cL \setminus \cL'$,
and traverse the chain of ancestors from $p$. Eventually, we must reach another short path $q$.
(If not, we have reached the root $r$ of $\pmax(T)$. Hence, $p$ is supported.)
Insert an edge from $p$ to $q$, so $q$ is the parent of $p$. This construction leads
to the shrub forest of $\cL \setminus \cL'$, where all the roots are supported short paths, 
and the remaining nodes are the unsupported short paths.

@@

We split the summation into tall, supported short, and unsupported short paths.
\begin{eqnarray*} 
\sum_{p \in \pmax(T)} \sum_{v \in p} \log |H_v| & = & \sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| + \sum_{p \in \cL'} \sum_{v \in p} \log |H_v| + \sum_{p \in \cH} \sum_{v \in p} \log |H_v|
%& = & \sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| + O(\sum_{p \in \cL'} |p|\log |H_p|) = \sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| + O(\sum_{p \in \cL'} |H_p|\log |H_p|)
\end{eqnarray*}
The last term can be bounded by $O(\sum_{p \in \pmax(T)} |p|\log |p|)$, by \Lem{pathBounds}.
The second term can be bounded by $O(\sum_{p \in \cL'} |H_p| \log |H_p|)$, again by \Lem{pathBounds}.
The following claim shows that this in turn is asymptotically bounded by the last term.

@@

\begin{proof} Let $\cU$ denote a connected component of the shrub forest.
We have $\sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| \allowbreak \leq \sum_{\cU} \sum_{p \in \cU} \sum_{v \in p} \log |H_v|$.
By \Lem{cu-root}, $|H_v| = O(|H_r|)$, where $r$ is the root of $\cU$, and furthermore, $\sum_{p \in \cU} |p| = O(|H_r|)$. 
We have $\sum_{p \in \cU} \sum_{v \in p} \log |H_v| = O((\log |H_r|) \sum_{p \in \cU} |p|) = O(|H_r|\log |H_r|)$.
We sum this over all $\cU$ in the shrub forest, and note that roots in the shrub forest are supported short paths.
\end{proof}

@@

We define a \emph{residue} $R_p$ for each $p \in \cU$. Suppose $p$ has children
$q_1, q_2, \ldots, q_k$ in $\cU$. Then $R_p = |\redH_p| - \sum_i |\redH_{q_i}|$. 
By definition, $|\redH_p| = \sum_{q \in \cU_p} R_q$. 
Note that $R_p$ can be negative. 
Now, define $R^+_p = \max(R_p,0)$, and set $W_p = \sum_{q \in \cU_p} R^+_q$, 
where $\cU_p$ is the subtree of $\cU$ rooted at $p$.
Observe that $W_p \geq |\redH_p|$. We also get an approximate converse.

@@

\begin{proof} We write $W_p - |\redH_p| = \sum_{q \in \cU_p} (R^+_q - R_q)$ $= -\sum_{q \in \cU_p: R_q < 0} R_q$.
Consider $q \in \cU_p$ and denote the children in $\cU_p$ by $q'_1, q'_2, \ldots$.
Note that $R_q$ is negative exactly when $|\redH_q| < \sum_i |\redH_{q'_i}|$.
Traverse $\pmax(T)$ from $q'_i$ to $q$.
Other than $q$, all other nodes encountered are in the tall support chain of $q'_i$ and hence
do not affect its reduced heap.
The vertices of $\redH_{q'_i}$ that are deleted
are exactly those present in the path $q$. 
Any vertex in $q$ can be deleted from at most two of the reduced heaps (of the children of $q$ in $\cU_p$), since 
theses reduced heaps do not have an ancestor-descendant relationship.
Therefore when $R_q$ is negative, it is at most by $2|q|$.
We sum over all $q$ to complete the proof.
\end{proof}
